---
title: "Additional Figures"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r, include = FALSE}
library(tidyverse)
library(httr)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
```


```{r Obtain API, include = FALSE}
data <- GET(
  url = "https://data.cdc.gov",
  path = "/resource/9bhg-hcku.json"
)
data <- content(data)
dat <- as.character(data)
```

```{r, include = FALSE}
covid_data <- read_csv("covid_death_data.csv")
```

```{r Web Scrapping,include = FALSE}
#

#Extract Gender
gender <- str_extract_all(dat, "sex = \"[:alpha:]+ [:alpha:]+|sex = \"[:alpha:]+")
gender <- str_remove_all(gender, "sex = \"")

#Extract state of death 
state <- str_extract_all(dat, "state = \"[:alpha:]+|state = \"[:alpha:]+ [:alpha:]+")
state <- str_remove_all(state, "state = \"")

#Extract age group 
age_group <- str_extract_all(dat, "age_group_new = \"[0-9]+-[0-9]+ [:alpha:]+|age_group_new = \"[:alpha:]+ [:alpha:]+|age_group_new = \"[:alpha:]+ [0-9]+ [:alpha:]+|age_group_new = \"[0-9]+ [:alpha:]+ [:alpha:]+ [:alpha:]+")
age_group <-str_remove_all(age_group, "age_group_new = \"")

#Covid-19 Deaths 
covid <- str_extract_all(dat," covid_19_deaths = \"[:alnum:]+")
covid <- str_remove_all(covid, " covid_19_deaths = \"")

#Pneumonia_deaths
pneumonia <- str_extract_all(dat, "pneumonia_deaths = \"[:alnum:]+")
pneumonia <- str_remove_all(pneumonia, "pneumonia_deaths = \"")

#Pneumonia and COVID Deaths 
both <- str_extract_all(dat, "pneumonia_and_covid_19_deaths = \"[:alnum:]+")
both <- str_remove_all(both,"pneumonia_and_covid_19_deaths = \"")

#Total_Deaths 
total <- str_extract_all(dat,"total_deaths = \"[:alnum:]+")
total <- str_remove_all(total, "total_deaths = \"")
```

```{r Create overall databse, message = FALSE,include = FALSE}
#Create and clean up COVID-19 death data frame 
database <- data.frame(
  Gender = gender,
  State = state,
  Age_Group = age_group,
  Covid_Deaths = covid,
  Pneumonia_Deaths = pneumonia,
  Covid_and_Pneumonia_Deaths = both,
  Total_Deaths = total
)

database %>% count(Gender)

knitr::kable(database)
```



```{r,include = FALSE}
#Group by independent variables. 
q1 <- data.frame(
  Gender = gender,
  State = state,
  Age_Group = age_group,
  Covid_Deaths = covid,
  Pneumonia_Deaths = pneumonia,
  Covid_and_Pneumonia_Deaths = both
)
q1 <- q1 %>% mutate(Covid_Deaths = as.numeric(Covid_Deaths))

q1 %>% count(Age_Group)
```



```{r Create Table for Age and COVID-19 Death,message = FALSE,include = FALSE}
q1 <- q1 %>% 
  filter(!(Covid_Deaths %in% NA)) %>% 
  filter(!(Age_Group %in% "All Ages"))%>%
  filter(!(Age_Group %in% "Under 1 year"))%>%
  filter(!(Age_Group %in% "1-4 years"))%>%
  filter(!(Age_Group %in% "0-17 years"))%>%
  filter(!(Age_Group %in% "18-29 years"))%>%
  filter(!(Age_Group %in% "30-49 years"))%>%
  filter(!(Age_Group %in% "50-64 years"))%>%
  filter(!(Age_Group %in% "85 years and over"))

Age<- q1 %>%
  filter((State %in% "United")) %>%
  filter(!(Gender %in% "All Sexes")) %>% 
  filter(!(Gender %in% "Unknown"))%>%
  group_by(Age_Group) %>% 
  summarise_at(vars(Covid_Deaths), list(Covid_Deaths=sum))

Age <- Age[c(5,1,2,3,4,6,7,8),]
sum(Age$Covid_Deaths, na.rm = TRUE)

knitr::kable(Age)  
```



```{r Create Table for Gender and COVID-19 Death,include = FALSE}
q1 <- q1 %>%
  filter(!(Gender %in% "All Sexes")) %>% 
  filter(!(Gender %in% "Unknown"))

gender_death <- q1 %>% 
  group_by(Gender)%>%
  summarise_at(vars(Covid_Deaths), list("Covid Death" = sum))

sum(gender_death$`Covid Death`)
```



```{r examine state, include = FALSE}
state_count <- covid_data %>% 
  filter(!(State %in% "United States")) %>%
  filter((Sex %in% "All Sexes"))%>%
  filter((`Age group` %in% "All Ages"))%>%
  group_by(State) %>%
  select(State,`COVID-19 Deaths`)

state_count <-state_count %>% 
  filter(!(State %in% "Puerto Rico"))

ny <- state_count[33, ]
nyc <- state_count[34, ]

state_count[33,2] <- ny$`COVID-19 Deaths`+nyc$`COVID-19 Deaths`

state_count <- state_count[-c(34),]
```
<br>



```{r,include = FALSE, warning = FALSE}
database <- database %>% mutate(Covid_Deaths = as.numeric(Covid_Deaths))
database <- database %>% mutate(Pneumonia_Deaths = as.numeric(Pneumonia_Deaths))
database <- database %>% mutate(Covid_and_Pneumonia_Deaths = as.numeric(Covid_and_Pneumonia_Deaths))
database <- database %>% mutate(Total_Deaths=as.numeric(Total_Deaths))

database <- database %>%
  filter(!(Covid_Deaths %in% NA)) %>% 
  filter((State %in% "United")) %>%
  filter(!(Age_Group %in% "All Ages"))%>%
  filter(!(Age_Group %in% "Under 1 year"))%>%
  filter(!(Age_Group %in% "1-4 years"))%>%
  filter(!(Age_Group %in% "0-17 years"))%>%
  filter(!(Age_Group %in% "18-29 years"))%>%
  filter(!(Age_Group %in% "30-49 years"))%>%
  filter(!(Age_Group %in% "50-64 years"))%>%
  filter(!(Age_Group %in% "85 years and over"))

q2a<- database %>%
  group_by(Age_Group) %>% 
  summarise_at(vars(Covid_Deaths), list(Covid_Deaths=sum))

q2b <- database %>% 
  filter(!(Pneumonia_Deaths %in% NA)) %>% 
  group_by(Age_Group) %>% 
  summarise_at(vars(Pneumonia_Deaths), list(Pneumonia_Deaths=sum))

q2c <- database %>% 
  filter(!(Covid_and_Pneumonia_Deaths %in% NA)) %>% 
  group_by(Age_Group) %>% 
  summarise_at(vars(Covid_and_Pneumonia_Deaths), list(Covid_and_Pneumonia_Deaths=sum))

q2 <-merge(Age,q2b, by = "Age_Group")
q2 <- merge(q2,q2c, by = "Age_Group")
  
q2 <- q2[c(5,1:4,6:8),]
```


### __Age__

The following table and figure analyze the relationship between age group and Covid-19 death: 

```{r Table for Age and COVID-19 Death,echo = FALSE}
knitr::kable(Age)  
```

The table above lists the number of Covid-19 deaths for each age group in the United States from February to August 2020. The numbers range from 35 deaths, for those between 5-14 years old, to 52,617 deaths, for those in between 75-84 years old.

<br>

```{r Create Figure for Age and COVID-19 Death,include = FALSE}
f1 <-Age %>% 
  filter(!(Age_Group %in% "5-14 years")) %>%
  ggplot(mapping = aes(x = Age_Group, y = Covid_Deaths))+
  geom_col()+
  labs(title = "Figure 1. Number of COVID-19 Deaths by Age Group in the United States,
February 2020 to August 2020")
```

```{r, echo = FALSE}
ggplotly(f1)
```


__Figure 1__ illustrates the number of Covid-19 deaths by age group in the United States from February to August 2020. The age group of 5-14 years was removed from the data set as the number of deaths due to Covid-19 was significantly less than the other age groups. There was 35 deaths from Covid-19 in the age group of 5-14, which comprised of <0.025% of the total Covid-19 death. 

According to __Figure 1__, the number of deaths due to Covid-19 increased for every age group. The older the patient, the greater the Covid-19 mortality rate. The increase in the number of deaths was particular pronounced after the age of 55, as the number of deaths seems to increase exponentially for each increase in age group. 

<br>

### __Gender__

The following table shows the difference in Covid-19 deaths by gender in the United States from February 2020 to August 2020. 

```{r, echo = FALSE}
knitr::kable(gender_death)
```

The number of males that have died due to Covid-19 is 114,291, while the number of females that have died due to Covid-19 is 75,203. These numbers are different from the total number of Covid-19 deaths calculated from the Covid-19 due to age distribution because different observations were omitted depending on the unknown or repetitive variables for each category. 

<br>

```{r, Gender and COVID-19, include = FALSE}
f2 <- gender_death %>% 
  ggplot(mapping = aes(x = Gender, y = `Covid Death`))+
  geom_col(width = 0.3)+
  labs(title = "Figure 2. Number of COVID-19 Deaths by Gender in the United States,
February to August 2020")
```

```{r, echo = FALSE}
ggplotly(f2)
```

__Figure 2__ illustrates the difference in the number of Covid-19 deaths in the United States by gender from February to August 2020. The cases in which the gender was unknown were removed from this figure. According to the data, there have been more cases of males dying due to Covid-19 than females. The ratio of males to females that have died due to COVID-19 is 1.520, indicating that the number of males that have died from Covid-19 is 1.520 times greater than the number of females that have died. 

<br>


```{r Figure of Covid-19 Death by Age and Gender, include = FALSE}
f3 <-q1 %>% 
  filter(!(Age_Group %in% "5-14 years")) %>%
  filter((State %in% "United")) %>%
  filter(!(Gender %in% "All Sexes")) %>% 
  filter(!(Gender %in% "Unknown"))%>%
  ggplot(mapping = aes(x = Age_Group, y = Covid_Deaths, fill = Gender))+
  geom_col(position = "dodge",colour = "black")+
  labs(title = "Figure 3.COVID-19 Deaths in the United States by Age Group and Gender", x = "Age Group", y = "Number of Covid-19 Deaths")
```

```{r, echo = FALSE, width = 1.5}
ggplotly(f3)
```

__Figure 3__ illustrates the number of Covid-19 deaths by age group and gender in the United States from February to August 2020. The gender distribution of the number of Covid-19 deaths for each age group mirrors that of the overall population of the United States. In each age group, the number of male deaths from Covid-19 are greater than that of females. 4

<br> 

### __State__

The following interactive table shows the number of deaths from Covid-19 for each state and the District of Columbia. The state of New York had a most deaths due to Covid-19 from February to August 2020, largely in part due to the large number of cases in New York City. The state of Wyoming had the least reported Covid-19 deaths during this time period. It is important to note, however, that this dataset does not included the population of each state, which can act as a counfounding factor for the number of deaths due to Covid-19. 

<br>

```{r, echo = FALSE}
datatable(state_count)
```
<br>

The following map shows the number of Covid-19 deaths for each state. Again, the number of Covid-19 deaths by state is the highest in New York. In general, the highest number of cases were concentrated around the New England area, Florida, Texas, and California. Many Midwestern states did not have many deaths due to Covid-19. As previously mentioned, differences in population, however, could effect the amount of covid-19 per state. 

```{r, include = FALSE}
state_count$state <- state_count$State

st_crosswalk <- tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
   bind_rows(tibble(state = "District of Columbia", abb = "DC"))
state_count <- left_join(state_count,st_crosswalk, by = "state")
state_count$hover <- with(state_count, paste(state, '<br>', "Deaths: ",`COVID-19 Deaths`))

set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

map <- plot_geo(state_count, locationmode = 'USA-states') %>%
  add_trace(
    z = ~`COVID-19 Deaths`, text = ~hover, locations = ~abb,
    color = ~`COVID-19 Deaths`, colorscale = 'magma'
  )

map <- map %>% colorbar(title = "Covid-19 Deaths by State")
map <- map %>% layout(
  title = paste('Covid-19 Death by State, from February 2020 to August 2020', '<br>(Hover for Value)'),
  geo = set_map_details)
```

```{r, echo = FALSE, fig.width = 10}
map
```

<br>

### __COVID-19 Deaths and Pneumonia Deaths by Age__

The following table shows the number of deaths for each condition (Covid-19, Pneumonia, Covid and Pneumonia) by age group. 

```{r Table for Q2, echo = FALSE}
knitr::kable(q2)
```

The number of deaths for each condition increased as the individual gets older, which is expected. The number of Covid deaths recorded does not include the number of deaths where both Covid-19 and pneumonia are found. 

<br>

```{r Figure of Age with Pnemonia and COVID Death, include = FALSE}
f4 <-q2 %>% 
  filter(!(Age_Group %in% "5-14 years")) %>% 
  ggplot(mapping = aes(x = Age_Group, y = Covid_and_Pneumonia_Deaths))+
  geom_col(color = "black")+
  labs(title = "Figure 4.Number of Deaths with Pneumonia and COVID-19", x = "Age Group", y = "Number of Covid-19 Deaths with Pneumonia")
```

```{r, echo = FALSE}
ggplotly(f4)
```


__Figure 4__ illustrates the number of deaths in which both Covid-19 and Pneumonia were involved. The number of cases with Covid-19 and pneumonia increase with age, and reaches the highest values at those between 75-84 years of age. 

<br>

```{r Figure of Proportoinal Deaths by Covid-19, include = FALSE}
q2 <- q2 %>% mutate(percent = Covid_and_Pneumonia_Deaths/(Covid_Deaths+Covid_and_Pneumonia_Deaths)*100)
f5 <- q2 %>% 
  filter(!(Age_Group %in% "5-14 years")) %>%
  filter(!(Covid_and_Pneumonia_Deaths %in% NA))%>%
  ggplot(mapping = aes(x = Age_Group,y = percent))+
  geom_col(colour = "black",width = 0.7)+
  labs(title = "Figure 5.Percentage of COVID-19 Deaths with Pneumonia by Age Group",x = "Age Group", y = "Proportion of COVID Deaths with Pneumonia")
```

```{r, echo = FALSE}
ggplotly(f5)
```

__Figure 5__ illustrates the percentage of deaths in which both Covid-19 and Pneumonia were present compared to the total number of deaths due to Covid-19. Throughout most age groups, the percentage of deaths in which both Covid-19 and Pneumonia are present account for around 40-50% of the total Covid-19 deaths. It stays relatively consistent throughout the age groups. 
